Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2020-06-02 source/sec_SGD.tex
sec 3 minor
sec 3.1:
1. remove the claim of first obtained linear speedup results in L152
2. rephrase comparison with related work.
3. typo in communication complexity of [19].

sec 3.2 remove N
change nonconvex to non-convex to be consistant accross the paper.

In this section, we provide convergence analyses of FedAvg with local
SGD updates in the general setting with heterogeneous data and partial
participation. We show that for strongly convex and smooth objectives,
the convergence of the optimality gap of averaged parameters across
devices is $\mathcal{O}(1/NT)$ where $N$ is the number of active
devices during each communication round, while for convex and smooth
objectives, the rate is $\mathcal{O}(1/\sqrt{NT})$.

@@

in the number of active workers during the communication round. To
our knowledge, this is the first result that explicitly demonstrates
linear speedup in the number of active workers in the setting with
both non i.i.d. data and partial participation. \\

@@

\textbf{Comparison with related works.} \cite{haddadpour2019convergence}
also proves a linear speedup $\mathcal{O}(1/NT)$ result for strongly
convex and smooth objectives in the heterogeneous data with partial
participation setting, with $\mathcal{O}(N^{1/3}T^{2/3})$ communication
complexity. However, their results are derived based on the more restrictive
bounded gradient diversity assumption. As discussed before, if $\frac{\sum_{k}p_{k}\|\nabla F_{k}(\mathbf{w})\|_{2}^{2}}{\|\sum_{k}p_{k}\nabla F_{k}(\mathbf{w})\|_{2}^{2}}$
is bounded, then $\nabla F_{k}(\mathbf{w}^{\ast})\equiv0$ for all
$k$. In this overparameterized setting, we show in Section~\ref{sec:overparameterized}
that the convergence is in fact exponential, thus improving on the
$\mathcal{O}(1/NT)$ rate in \cite{haddadpour2019convergence} with

@@

\subsection{Convex Smooth Objectives}
Next we provide linear speedup analyses of FedAvg with convex and
smooth objectives and show that the optimality gap is $\mathcal{O}(1/\sqrt{NT})$
where $N$ is the number of participating devices. This result complements
the strongly convex case in the previous part, as well as the nonconvex
smooth setting in \cite{jiang2018linear,yu2019parallel,haddadpour2019convergence},
where $\mathcal{O}(1/\sqrt{NT})$ results are given in terms of averaged
gradient norm. 
\begin{theorem}
	\label{thm:SGD_cvx}With full device participation and constant learning
	rate $\alpha_{t}=\mathcal{O}(\sqrt{\frac{N}{T}})$, 
	\begin{align*}
	\min_{t\leq T}F(\overline{\mathbf{w}}_{t})-F(\mathbf{w}^{\ast}) & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{NT}}+\frac{NE^{2}LG^{2}}{T}\right)
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round and learning rate $\alpha_{t}=\mathcal{O}(\sqrt{\frac{K}{T}})$,
	\begin{align*}
	\min_{t\leq T}F(\overline{\mathbf{w}}_{t})-F(\mathbf{w}^{\ast}) & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{KT}}+\frac{E^{2}G^{2}}{\sqrt{KT}}+\frac{KE^{2}LG^{2}}{T}\right)
	\end{align*}
\end{theorem}
%
\textbf{Choice of $E$ and linear speedup. }With full participation,
as long as $E=\mathcal{O}(T^{1/4}/N^{3/4})$, the convergence
rate is $\mathcal{O}(1/\sqrt{NT})$ with $\mathcal{O}(N^{3/4}T^{3/4})$
communication rounds. In the partial participation setting, $E$ must
be $O(1)$ in order to achieve linear speedup of $\mathcal{O}(1/\sqrt{KT})$.
Our result again demonstrates the difference in communication complexities
between full and partial participation, and is to our knowledge the
first result on linear speedup in the general federated learning setting
with both heterogeneous data and partial participation. \\
\textbf{Learning rate. }The learning rate now depends on the final
horizon $T$ of the convergence statement, whereas in the strongly
convex case the learning rate decays as $\mathcal{O}(1/t)$. Such
a requirement $\alpha_{t}=\mathcal{O}(\sqrt{N/T})$ is also present
in the works \cite{haddadpour2019convergence,yu2019parallel} on nonconvex
problems with $\mathcal{O}(1/\sqrt{NT})$ linear speedup convergence results.

>>>>

In this section, we provide convergence analyses of FedAvg with local
SGD updates in the general setting with heterogeneous data and partial
participation. We show that for strongly convex and smooth objectives,
the convergence of the optimality gap of averaged parameters across
devices is $\mathcal{O}(1/NT)$, while for convex and smooth
objectives, the rate is $\mathcal{O}(1/\sqrt{NT})$. 
The proof of FedAvg is deferred in Appendix Section~\ref{sec:app:fedavg}.

@@

in the number of active workers during the communication round. \\
% To our knowledge, this is the first result that explicitly demonstrates
% linear speedup in the number of active workers in the setting with
% both non i.i.d. data and partial participation. \\

@@

\textbf{Comparison with related works.} To better understand the significance of obtained bound, we compare 
our rates to the best-known results in the same setting. 
In~\cite{haddadpour2019convergence}, the authors prove a linear speedup $\mathcal{O}(1/NT)$ result for strongly
convex and smooth objectives, with $\mathcal{O}(N^{1/3}T^{1/3})$ communication complexity when considering i.i.d. data. 
However, their results build upon the bounded gradient diversity assumption, which leads to a more restricted overparameterized setting (see discussions in Section~\ref{sec:assumptions}). In this overparameterized setting, we prove a geometric convergence rate (see Section~\ref{sec:overparameterized}), 
thus improving on the rate in \cite{haddadpour2019convergence} with

@@

\subsection{Convex Smooth Objectives}
Next we provide linear speedup analyses of FedAvg with convex and
smooth objectives and show that the optimality gap is $\mathcal{O}(1/\sqrt{NT})$. 
This result complements the strongly convex case in the previous part, as well as the non-convex
smooth setting in \cite{jiang2018linear,yu2019parallel,haddadpour2019convergence},
where $\mathcal{O}(1/\sqrt{NT})$ results are given in terms of averaged
gradient norm. 
\begin{theorem}
	\label{thm:SGD_cvx}With full device participation and constant learning
	rate $\alpha_{t}=\mathcal{O}(\sqrt{\frac{N}{T}})$, 
	\begin{align*}
	\min_{t\leq T}F(\overline{\mathbf{w}}_{t})-F(\mathbf{w}^{\ast}) & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{NT}}+\frac{NE^{2}LG^{2}}{T}\right)
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round and learning rate $\alpha_{t}=\mathcal{O}(\sqrt{\frac{K}{T}})$,
	\begin{align*}
	\min_{t\leq T}F(\overline{\mathbf{w}}_{t})-F(\mathbf{w}^{\ast}) & =\mathcal{O}\left(\frac{\nu_{\max}^{2}\sigma^{2}}{\sqrt{KT}}+\frac{E^{2}G^{2}}{\sqrt{KT}}+\frac{KE^{2}LG^{2}}{T}\right)
	\end{align*}
\end{theorem}
%
\textbf{Choice of $E$ and linear speedup. }With full participation,
as long as $E=\mathcal{O}(T^{1/4}/N^{3/4})$, the convergence
rate is $\mathcal{O}(1/\sqrt{NT})$ with $\mathcal{O}(N^{3/4}T^{3/4})$
communication rounds. In the partial participation setting, $E$ must
be $O(1)$ in order to achieve linear speedup of $\mathcal{O}(1/\sqrt{KT})$.
Our result again demonstrates the difference in communication complexities
between full and partial participation, and is to our knowledge the
first result on linear speedup in the general federated learning setting
with both heterogeneous data and partial participation. \\
\textbf{Learning rate. }The learning rate now depends on the final
horizon $T$ of the convergence statement, whereas in the strongly
convex case the learning rate decays as $\mathcal{O}(1/t)$. Such
a requirement $\alpha_{t}=\mathcal{O}(\sqrt{N/T})$ is also present
in the works \cite{haddadpour2019convergence,yu2019parallel} on non-convex
problems with $\mathcal{O}(1/\sqrt{NT})$ linear speedup convergence results.
